---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```
## Loading and preprocessing the data
```{r load-data}
# Load the dataset
activity <- read.csv("activity.csv")

# Preview the dataset
head(activity)
```
## What is mean total number of steps taken per day?

```{r total-steps}
# Calculate total steps per day
total_steps_per_day <- activity %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = TRUE))

# Histogram of total steps per day
ggplot(total_steps_per_day, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "skyblue", color = "black") +
  labs(title = "Total Steps per Day", x = "Total Steps", y = "Frequency")


# Mean and median of total steps
mean_steps <- mean(total_steps_per_day$total_steps)
median_steps <- median(total_steps_per_day$total_steps)

mean_steps
median_steps
```

## What is the average daily activity pattern?

```{r daily-pattern}
# Average steps for each interval
average_steps_per_interval <- activity %>%
  group_by(interval) %>%
  summarize(mean_steps = mean(steps, na.rm = TRUE))

# Time series plot of average steps

ggplot(average_steps_per_interval, aes(x = interval, y = mean_steps)) +
  geom_line(color = "blue") +
  labs(title = "Average Daily Activity Pattern", x = "5-minute Interval", y = "Average Steps")

# 5-minute interval with maximum steps
max_interval <- average_steps_per_interval[which.max(average_steps_per_interval$mean_steps),]
max_interval
```
## Imputing missing values

```{r imputing}
# Total number of missing values
total_missing <- sum(is.na(activity$steps))
total_missing

# Strategy: fill missing values with mean of the corresponding interval
impute_mean <- function(steps, interval) {
  if (is.na(steps)) {
    return(average_steps_per_interval$mean_steps[average_steps_per_interval$interval == interval])
  } else {
    return(steps)
  }
}

# Create a new dataset with missing values filled
activity_imputed <- activity
activity_imputed$steps <- mapply(impute_mean, activity_imputed$steps, activity_imputed$interval)

# Histogram of total steps per day (after imputation)
total_steps_per_day_imputed <- activity_imputed %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))

ggplot(total_steps_per_day_imputed, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "lightgreen", color = "black") +
  labs(title = "Total Steps per Day (Imputed)", x = "Total Steps", y = "Frequency")

# Mean and median after imputation
mean_steps_imputed <- mean(total_steps_per_day_imputed$total_steps)
median_steps_imputed <- median(total_steps_per_day_imputed$total_steps)

mean_steps_imputed
median_steps_imputed
```
## Are there differences in activity patterns between weekdays and weekends?

```{r weekday-weekend}
# Create a factor variable for day type
activity_imputed$date <- as.Date(activity_imputed$date)
activity_imputed$day_type <- ifelse(weekdays(activity_imputed$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")

# Average steps by interval and day type
average_steps_daytype <- activity_imputed %>%
  group_by(interval, day_type) %>%
  summarize(mean_steps = mean(steps))

# Panel plot

ggplot(average_steps_daytype, aes(x = interval, y = mean_steps)) +
  geom_line(color = "purple") +
  facet_wrap(~day_type, ncol = 1) +
  labs(title = "Average Steps by Interval: Weekday vs Weekend", x = "5-minute Interval", y = "Average Steps")
